Measuring contrast processing in the visual system using SSVEP

### 2024-04-04 review.qmd
Fixed some references

---
title: "Measuring contrast processing in the visual system using SSVEP (and VEP?)"
format: pdf
bibliography: WadeBaker_Review.bib
csl: visual-neuroscience.csl
execute:
  echo: false
---

## Baker, Wade: Review: Vis Neurosci

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] (Hillebrand?). The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well. [McKeefry, Sutter, Baseler, DiRusso...].

Contrast is relatively simple to define: typically, contrast is specified as the percentage deviation of a uniform stimulus from the background. So, for example, a disk of 100 units of cone activation ($I_{\mathrm{stim}}$) surrounded by a 'background' of 50 units of activation ($I_{\mathrm{background}}$) has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the @Michelson1927 definition of contrast is specified by the maximum and minimum excursions from the mean: $${{\frac {I_{\mathrm {stimmax} }-I_{\mathrm {stimmin} }}{I_{\mathrm {stimmax} }+I_{\mathrm {stimmin} }}}.} $$ These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance [LenniePokornySmith]) and also to definitions based on cone excitations [@Derrington1984, @MacLeod1979] which are more common in work on chromatic processing.

@@

Here, we show the effect of adding a second contrast component: in general, the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan]

@@

\begin{equation}
R_{max} = \frac{C_{in}^n}{C_{50}^n + C_{in}^n},
\end{equation}

where ${R_{max}}$ describes the maximum response level, $C_{in}$ is the input contrast, $C_{50}$ is the 'semi-saturation constant' (the point at which the reponse is at half-maximum) and *n* controls the steepness of the curve.

@@

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. Candy and Norcia (in infants) 2001, Busse et al2008, Tsai/Norcia/Wade 2010s, Then lots of Baker stuff including the CerebCort paper. : Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990]

@@

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [@Norcia, Tyler, Braddick] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

Because SSVEP response at detection threshold are very small, estimating this threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis. This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@DaleNorciaTyler86].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler, Nakayama etc 1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [Tyler, Norcia etc: 1985, 1990 etc].

This general approach has revealed much about the development of visual abilities in infants [@braddick, @atkinson]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@dobsonVisualAcuityHuman1978; @norciaSpatialFrequencySweep1985, ] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months while spatial acuity as measured by SSVEP reaches adult levels more slowly but near-adult levels are recorded around 1 year [@NorciaTyler1985, @HamerVisRed1989] compared to  around 6-7 years with behavioural measures [@Ellemberg vis res 1999, @Atkinson]. At least some of this difference is likely due to the relatively objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking which require careful measurement of the infant's gaze direction although it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@de1982maturation].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section.[@Candy 2001, @PeiNorcia 2017]. Although contrast gain control is measurable in infants as young as six weeks old (@Candy2001, @MorroneBurr1986, @Skoczenski & Norcia, 1998), development appears to be slower with adult levels being reached in approximately 11 years. [@PeiNorcia].

@@

The sweep VEP (technically, the sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. Because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004 Diva vs Enfant, @Hoffman, @Bach and @Farmer].

>>>>

---
title: "Measuring contrast processing in the visual system using SSVEP"
format: pdf
bibliography: WadeBaker_Review.bib
csl: visual-neuroscience.csl
execute:
  echo: false
---

## Wade, A.R. & Baker, D.H.: Review for Vis Neurosci

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] (@Hillebrand?). The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@McKeefry1996; @Baseler1997; @Di-Russo2001].

Contrast is relatively simple to define: typically, contrast is specified as the percentage deviation of a uniform stimulus from the background. So, for example, a disk of 100 units of cone activation ($I_{\mathrm{stim}}$) surrounded by a 'background' of 50 units of activation ($I_{\mathrm{background}}$) has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the @Michelson1927 definition of contrast is specified by the maximum and minimum excursions from the mean:

\begin{equation}
\frac {I_{\mathrm{stimmax} }-I_{\mathrm{stimmin} }}{I_{\mathrm{stimmax} }+I_{\mathrm{stimmin} }}.
\end{equation}

These contrast definitions are appropriate both to photometric measures of stimulus contrast [for example, luminance; @Lennie1993] and also to definitions based on cone excitations [@Derrington1984, @MacLeod1979] which are more common in work on chromatic processing.

@@

Here, we show the effect of adding a second contrast component: in general, the two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988]

@@

\begin{equation}
R_{\mathrm{max}} = \frac{C_{\mathrm{in}}^n}{C_{50}^n + C_{\mathrm{in}}^n},
\end{equation}

where ${R_{\mathrm{max}}}$ describes the maximum response level, $C_{\mathrm{in}}$ is the input contrast, $C_{50}$ is the 'semi-saturation constant' (the point at which the reponse is at half-maximum) and *n* controls the steepness of the curve.

@@

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. @Candy2001; @Busse2009; @Tsai2012; @Baker2017 Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990]

@@

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [@Norcia1990; Tyler, Braddick] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

Because SSVEP response at detection threshold are very small, estimating this threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis. This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler, Nakayama etc 1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [Tyler, @Norcia1985a; @Norcia1990].

This general approach has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around 1 year [@Norcia1985b, @HamerVisRed1989] compared to  around 6-7 years with behavioural measures [@Ellemberg1999; @Atkinson]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section.[@Candy2001; @Pei2017]. Although contrast gain control is measurable in infants as young as six weeks old [@Candy2001; @Morrone1986; @Skoczenski1998], its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

@@
